---
title: "Clean Policy Data"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

```{r, load data markdown}
rmarkdown::render("load_data.Rmd")
```

## Prepare Mask Policy Data
#Calculate the longest number of days that any state had a mask mandate policy, which would be whichever state implemented it first and then up to the most recent date a state ended their mask mandate
-Column 'Public face mask mandate' = beginning date
-Column 'End face mask mandate' = ending date
```{r, Calculate mask mandate date range}
mask_policy <- mask_policy[-52,]

first_mask <- mask_policy %>% filter(mask_policy$Public.face.mask.mandate != 0)
last_mask <- mask_policy %>% filter(mask_policy$End.face.mask.mandate != 0)

first_mask$Public.face.mask.mandate <- as.Date(first_mask$Public.face.mask.mandate, format = "%m/%d/%Y")
last_mask$End.face.mask.mandate <- as.Date(last_mask$End.face.mask.mandate, format = "%m/%d/%Y")

mask_begin <- min(first_mask$Public.face.mask.mandate)
mask_end <- max(last_mask$End.face.mask.mandate)

mask_range <- mask_end - mask_begin

mask_range <- as.numeric(mask_range, units = "days")
```

#Add a column for the number of days each state had/has a mask mandate policy
```{r, Calculate number of days each state had a mask policy}
mask_policy$Public.face.mask.mandate <- as.Date(mask_policy$Public.face.mask.mandate, format = "%m/%d/%Y")
mask_policy$End.face.mask.mandate <- as.Date(mask_policy$End.face.mask.mandate, format = "%m/%d/%Y")

mask_policy <- mask_policy %>% mutate(days_mask = case_when(
  is.na(Public.face.mask.mandate) == TRUE ~ 0,
  is.na(Public.face.mask.mandate) == FALSE & is.na(End.face.mask.mandate) == TRUE ~ as.numeric((mask_end - Public.face.mask.mandate), units = "days"),
  TRUE ~ as.numeric((End.face.mask.mandate - Public.face.mask.mandate), units = "days")))
```

#Add a column to the mask_policy data frame to show the ratio of days each state mandate masks
```{r, Calculate ratio of policy in place to the total number of days masks have been required}
mask_policy$mask_ratio <- mask_policy$days_mask / mask_range
```

## Prepare Stay-at-Home Policy Data
#Calculate the longest number of days that any state had a stay-at-home mandate policy, which would be whichever state implemented it first and then up to the most recent date a state ended their stay-at-home mandate
-Column 'Stay at home/shelter in place' = beginning date
-Column 'End stay at home/shelter in place ' = ending date
```{r, Calculate stay-at-home mandate date range}
stayathome_policy <- stayathome_policy[-52,]

first_stay <- stayathome_policy %>% filter(stayathome_policy$Stay.at.home.shelter.in.place != 0)
last_stay <- stayathome_policy %>% filter(stayathome_policy$End.stay.at.home.shelter.in.place != 0)

first_stay$Stay.at.home.shelter.in.place <- as.Date(first_stay$Stay.at.home.shelter.in.place, format = "%m/%d/%Y")
last_stay$End.stay.at.home.shelter.in.place <- as.Date(last_stay$End.stay.at.home.shelter.in.place, format = "%m/%d/%Y")

home_begin <- min(first_stay$Stay.at.home.shelter.in.place)
home_end <- max(last_stay$End.stay.at.home.shelter.in.place)

home_range <- home_end - home_begin

home_range <- as.numeric(home_range, units = "days")
```

#Add a column for the number of days each state had/has a stay-at-home policy
```{r, Calculate number of days each state had a mask policy}
stayathome_policy$Stay.at.home.shelter.in.place <- as.Date(stayathome_policy$Stay.at.home.shelter.in.place, format = "%m/%d/%Y")
stayathome_policy$End.stay.at.home.shelter.in.place <- as.Date(stayathome_policy$End.stay.at.home.shelter.in.place, format = "%m/%d/%Y")

stayathome_policy <- stayathome_policy %>% mutate(days_home = case_when(
  is.na(Stay.at.home.shelter.in.place) == TRUE ~ 0,
  is.na(Stay.at.home.shelter.in.place) == FALSE & is.na(End.stay.at.home.shelter.in.place) == TRUE ~ as.numeric((home_end - Stay.at.home.shelter.in.place), units = "days"),
  TRUE ~ as.numeric((End.stay.at.home.shelter.in.place - Stay.at.home.shelter.in.place), units = "days")))
```

#Add a column to the stayathome_policy data frame to show the ratio of days each state mandate to stay at home 
```{r, Calculate ratio of policy in place to the total number of days masks have been required}
stayathome_policy$home_ratio <- stayathome_policy$days_home / home_range
```
