---
title: "Causes of Varying Death Rates in the U.S."
subtitle: "W203 Lab 2"
author: "Jenny Conde, Spencer Hong, Kevin Lu, Spencer Song"
output: pdf_document
---

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(car)
library(lmtest)
library(sandwich)
library(stargazer)
library(tidyverse)
library(magrittr)
library(ggplot2)
library(dplyr)
library(patchwork)
```

# Introduction

<!-- Your introduction should present a research question and explain the concept that you're attempting to measure and how it will be operationalized. This section should pave the way for the body of the report, preparing the reader to understand why the models are constructed the way that they are. It is not enough to simply say "We are looking for policies that help against COVID."  Your introduction must do work for you, focusing the reader on a specific measurement goal, making them care about it, and propelling the narrative forward. This is also good time to put your work into context, discuss cross-cutting issues, and assess the overall appropriateness of the data.

(From Rubric) Is the introduction clear? Is the research question specific and well defined? Does the introduction motivate a specific concept to be measured and explain how it will be operationalized. Does it do a good job of preparing the reader to understand the model specifications?

TODO: jenny conde
Driving question: What causes higher death rates in some states versus others?
  What factors impact ___
Variable categories: Health, policy, socioeconomic, other -->

The COVID-19 pandemic has overturned nearly all aspects of normal life over the past year. The U.S. has particularly suffered, with 550,000 COVID-19 deaths, more than any other nation in the world (https://www.nytimes.com/interactive/2020/us/coronavirus-us-cases.html). However, not all areas of the U.S. have been equally affected. While some states have seen relatively few cases and deaths, others have been significantly impacted, seeing overwhelmed healthcare systems and lockdowns. This motivates our primary research question: What causes higher death rates in some states versus others? What are the primary factors driving higher death rates across states?

To answer this question, we analyze the cumulative COVID-19 death rate from ____ through ____. We calculate the death rate as the number of deaths per positive cases. We operationalize several categories of factors that likely have a causal relationship with death rate based on our research from media articles, publications, and government sources. These categories include pre-existing health conditions, enacted policy, socioeconomic characteristics. We include a few miscellaneous features not captured in these broad categories, including concentration of ICU beds.

With this background knowledge about factors making people more prone to catching a severe case of COVID-19, we can perform a causal analysis to determine how prevalent these factors are at a state-level. Our causal analysis consists of three models. Our first model contains the variables we believe have the largest causal impact on death rate. Our second model builds on this to include more of our dataset, and finally, our third model includes all data we collected.

### Intermediate section: data collection/aggregation/cleaning

TODO: spencer song
- explain dropping of Hawaii, Alaska, D.C.
- explain how we define "death rate" (i.e. deaths/cases)--why is this operationalization more relevant than alternatives (ex. deaths per capita) -- Jenny explains this in the intro a bit but this operationalization probably merits further explanation

<!-- 
Note from Jenny: I originally wrote this section for the intro, but it's probably too detailed?? Feel free to use snippets of this.

Our operationalized pre-existing health conditions were largely informed by the Center for Disease Control's list of medical conditions (https://www.cdc.gov/coronavirus/2019-ncov/need-extra-precautions/people-with-medical-conditions.html) that make adults more likely to get severely ill from COVID-19. We focused our policy data on mask requirements and stay at home orders. Individuals' socioeconomic status has also been known to make them more vulnerable to COVID-19 (https://www.mckinsey.com/featured-insights/coronavirus-leading-through-the-crisis/charting-the-path-to-the-next-normal/socioeconomic-vulnerability-increases-the-risk-of-dying-from-covid-19#:~:text=Socioeconomic%20vulnerability%20increases%20the%20risk%20of%20dying%20from%20COVID%2D19,-COVID%2D19%20Inequality&text=September%2010%2C%202020%20People%20who,coronavirus%20than%20the%20general%20population.), so we include features like unemployment, income, and crowding. Finally, we analyze a few additional factors: percent of people over age 65 and concentration of ICU beds, which have also been known to have an impact on COVID-19. 

Also, maybe we should make a table of our data, showing how they're grouped into categories and the sources?

-->

1. \textbf{Explanatory Variable}

\vspace{-5pt}

> Fatality Rate [(NY Times)](https://github.com/nytimes/covid-19-data/blob/master/live/us-counties.csv): calculated as the ratio of cumulative deaths to cumulative cases

2. \textbf{Health Variables}

\vspace{-5pt}

> % Adults with Diabetes [(CDC)](https://gis.cdc.gov/grasp/diabetes/DiabetesAtlas.html#)

\vspace{-5pt}

> % Deaths from Stroke [(CDC)](https://nccd.cdc.gov/DHDSPAtlas/?state=County)

\vspace{-5pt}

> % Deaths from Heart Disease [(CDC)](https://nccd.cdc.gov/DHDSPAtlas/?state=County)

\vspace{-5pt}

> % Adults who Smoke [(CHR)](https://www.countyhealthrankings.org/explore-health-rankings/rankings-data-documentation)

\vspace{-5pt}

> % Deaths from Chronic Respiratory Disease [(GHDx)](http://ghdx.healthdata.org/record/ihme-data/united-states-chronic-respiratory-disease-mortality-rates-county-1980-2014)

\vspace{-5pt}

> % Deaths from Asthma [(GHDx)](http://ghdx.healthdata.org/record/ihme-data/united-states-chronic-respiratory-disease-mortality-rates-county-1980-2014)

\vspace{-5pt}

> % Adults with Obesity [(CHR)](https://www.countyhealthrankings.org/explore-health-rankings/rankings-data-documentation)



3. \textbf{Policy Variables}

\vspace{-5pt}

> Mask Ratio [(W203)](https://docs.google.com/spreadsheets/d/1zu9qEWI8PsOI_i8nI_S29HDGHlIp2lfVMsGxpQ5tvAQ/edit#gid=1348247662): calculated as the proportion of days each state had a mask mandate between April 8, 2020, and April 4, 2021

\vspace{-5pt}

> Stay at Home Ratio [(W203)](https://docs.google.com/spreadsheets/d/1zu9qEWI8PsOI_i8nI_S29HDGHlIp2lfVMsGxpQ5tvAQ/edit#gid=1348247662): calculated as the proportion of days each state had a stay at home mandate between March 19, 2020, and January 25, 2021

4. \textbf{Socioeconomic Variables}

\vspace{-5pt}

> % Unemployed [(CDC)](https://www.atsdr.cdc.gov/placeandhealth/svi/data_documentation_download.html)

\vspace{-5pt}

> % Minority [(CDC)](https://www.atsdr.cdc.gov/placeandhealth/svi/data_documentation_download.html)

\vspace{-5pt}

> % Crowding [(CDC)](https://www.atsdr.cdc.gov/placeandhealth/svi/data_documentation_download.html)

\vspace{-5pt}

> % Adults Living in Multi-Unit Dwellings [(CDC)](https://www.atsdr.cdc.gov/placeandhealth/svi/data_documentation_download.html)

\vspace{-5pt}

> % Adults Living in Poverty [(CDC)](https://www.atsdr.cdc.gov/placeandhealth/svi/data_documentation_download.html)

\vspace{-5pt}

> Median (?) Household Income

5. \textbf{Miscellaneous Variables}

\vspace{-5pt}

> % Adults Over Age 65 [(CHR)](https://www.countyhealthrankings.org/explore-health-rankings/rankings-data-documentation)

\vspace{-5pt}

> ICU Beds per 10,000 people [(KFF)](https://www.kff.org/other/state-indicator/icu-beds/?currentTimeframe=0&sortModel=%7B%22colId%22:%22Location%22,%22sort%22:%22asc%22%7D)

### 2. A Model Building Process -- EVERYONE go through your vars and determine how your EDA should impact model building (All, led by Spencer Hong)

<!--You will next build a set of models to investigate your research question, documenting your decisions. Here are some things to keep in mind during your model building process:

1. *What do you want to measure*? Make sure you identify one key variable (possibly more in rare cases) that will allow you to derive conclusions relevant to your research question, and include this variables in all model specifications.

2. Is your modeling goal one of description or explanation?-

3. What [covariates](https://en.wikipedia.org/wiki/Dependent_and_independent_variables#Statistics_synonyms) help you achieve your modeling goals? What covariates are problematic, either due to *collinearity*, or because they will absorb some of a causal effect you want to measure?

4. What *transformations*, if any, should you apply to each variable? These transformations might reveal linearities in the data, make your results relevant, or help you meet model assumptions.

5. Are your choices supported by exploratory data analysis (*EDA*)? You will likely start with some general EDA to *detect anomalies* (missing values, top-coded variables, etc.). From then on, your EDA should be interspersed with your model building. Use visual tools to *guide* your decisions. You can also leverage statistical *tests* to help assess whether variables, or groups of variables, are improving model fit.
  - remove mask policy?-->

1. What do you want to measure?
The team identified fatality rates as the key variable to be measured as we look to understand the relationship between fatality rates and several chosen variables that fall within the health, policy, or socioeconomic realms.

2. Is your modeling goal one of description or explanation?
Our model is an explanatory one as we not only describe the COVID data and trends but also provide causal relationships between each chosen indepedent and dependent variables. We plan on explaining causes in the changes in fatality rates.

3. What [covariates](https://en.wikipedia.org/wiki/Dependent_and_independent_variables#Statistics_synonyms) help you achieve your modeling goals? What covariates are problematic, either due to *collinearity*, or because they will absorb some of a causal effect you want to measure?

Health: We selected seven different health covariates  Diabetes_Diagnosed_Perc, Stroke_Death_Perc, Heart_Disease_Death_Perc,
Smoker_Perc, Chronic_Respiratory_Disease_Death_Perc, Asthma_Deaths_Perc, Obesity_Perc

Policy: The covariates for both mask mandate and stay-at-home policy durations are of concern as they both show a positive linear relationship with fatality rates.

Socioeconomic:

Miscellaneous:

4. What *transformations*, if any, should you apply to each variable? These transformations might reveal linearities in the data, make your results relevant, or help you meet model assumptions.

Health:

Policy: The covariates for both mask mandate and stay-at-home policy duratiions are of concern as they both show a positive linear relationship with fatality rates.

Socioeconomic:

Miscellaneous:

<!--5. Are your choices supported by exploratory data analysis (*EDA*)? You will likely start with some general EDA to *detect anomalies* (missing values, top-coded variables, etc.). From then on, your EDA should be interspersed with your model building. Use visual tools to *guide* your decisions. You can also leverage statistical *tests* to help assess whether variables, or groups of variables, are improving model fit.
  - remove mask policy?-->
  
5. Are your choices supported by exploratory data analysis (*EDA*)?

TODO: each person individually does their own EDA and write-up
TODO: EDA for all 3 model types and reasoning for specific applied transformations
  - remove NA and check for reasoning with why they existed (make sure not all from one state, etc.)
  - scatter, hists, bar?, box 
  - variable summary, outlier removal? / reasoning
  - transformation logic (transformations to use,etc. include reasoning)

At the same time, it is important to remember that you are not trying to create one perfect model. You will create several specifications, giving the reader a sense of how robust (or sensitive) your results are to modeling choices, and to show that you're not just cherry-picking the specification that leads to the largest effects.

At a minimum, you should include the following three specifications:

1. **Limited Model**: A model that includes *only the key variable* you want to measure and nothing (or almost nothing) else. This variables might be transformed, as determined by your EDA, but the model should include the absolute minimum number of covariates (perhaps one, or at most two-three, covariates if they are so crucial that it would be unreasonable to omit them). 
1. **Model Two**: A model that includes *key explanatory variables and covariates that you believe advance your modeling* goals without introducing too much multicollinearity or causing other issues. This model should strike a balance between accuracy and parsimony and reflect your best understanding of the relationships among key variables.
1. **Model Three**: A model that includes the *previous covariates, and many other covariates*, erring on the side of inclusion. A key purpose of this model is to evaluate how parameters of interest change (if at all) when additional, potentially colinear variables are included in the model specification.
TODO: data summary paragraph / report
  TODO: anomaly report


TODO: Spencer Hong

TODO: assess addition of each category of data from model 1 to model 2? (how does adding all ___ data to limited model impact the model?)
  TODO: everyone can try this for their section. Take the limited model, and then add vars for model 2 from your section to the linear model. Analyze the difference and see how your variables impacted the model.
  
```{r, echo = FALSE, message = FALSE, warning = FALSE}
df <- read_csv('FinalDF.csv')
df <- df[-c(2, 9, 12), ] # remove Alaska, Hawaii, D.C.
```


TODO: Limited:
  1. health, 1. policy, 1. socioeconomic
  health var: chronic respitory disease / smoking
  policy var: mask policy ratio
  socioeconomic: poverty / minority
  general: over 65%
  
```{r}
model_limited <- lm(Fatality_Rate ~ mask_ratio + Senior_Perc + Poverty_Perc, data = df)
coeftest(model_limited)
```

  
TODO: Model 2: do NOT include these:
  policy - include all
  demographics - ICU data
  socioeconomic - multi unit dwelling, income/unemployment/one or the other
  health - asthma death, obesity

```{r}
model_goldilocks <- lm(Fatality_Rate ~ mask_ratio + home_ratio + Diabetes_Diagnosed_Perc + Stroke_Death_Perc+ Heart_Disease_Death_Perc + Smoker_Perc + Chronic_Respiratory_Disease_Death_Perc + Senior_Perc + Poverty_Perc + log(Minority_Percentage) + Unemployment_Perc  + Crowding_Perc, data = df)
coeftest(model_goldilocks)
```

TODO: Model 3:
  everything we have
  
```{r}
model_big <- lm(Fatality_Rate ~ mask_ratio + home_ratio + Diabetes_Diagnosed_Perc + Stroke_Death_Perc+ Heart_Disease_Death_Perc + Smoker_Perc + Chronic_Respiratory_Disease_Death_Perc + Asthma_Deaths_Perc +  Obesity_Perc + Senior_Perc + Poverty_Perc + log(Minority_Percentage) + Unemployment_Perc  + Crowding_Perc + Multi_Unit_Perc + HouseholdIncome + ICUBedsPer10000, data = df)
coeftest(model_big)
```


Although the models have different emphases, each one must still be a reasonable choice given your modeling goals.  The idea is to choose models that encircle the space of reasonable modeling choices, and to give an overall understanding of how these choices impact results.

### 3. A Regression Table -- Kevin

You should display all of your model specifications in a regression table, using a package like [`stargazer`](https://cran.r-project.org/web/packages/stargazer/vignettes/stargazer.pdf) to format your output. It should be easy for the reader to find the coefficients that represent key effects at the top of the regression table, and scan horizontally to see how they change from specification to specification. Make sure that you display the most appropriate standard errors in your table, along with significance stars.

In your text, comment on both *statistical significance and practical significance*. You may want to include statistical tests besides the standard t-tests for regression coefficients.

TODO: run packages on data
TODO: comment on statistical and practical significance

### 4. Limitations of your Model -- All, led by Jenny

As a team, evaluate all of the CLM assumptions that must hold for your model. However, do not report an exhaustive examination all 5 CLM assumption. Instead, bring forward only those assumptions that you think pose significant problems for your analysis. For each problem that you identify, describe the statistical consequences. If you are able to identify any strategies to mitigate the consequences, explain these strategies. 

Note that you may need to change your model specifications in response to violations of the CLM. 

TODO: CLM Assumptions and reasoning for violations
  TODO: how those violations should change our model specifications
TODO: strategies to mitigate violations

TODO: causality concerns / collinearity
  - check plots of var v. var
  - correlation matrix
  - draw causal theories?

- IID

- Linear conditional expectation
###> To assess whether there is a linear conditional expectation, we've learned to look at the predicted vs. residuals of the model.
```{r}
df %>% 
  mutate(
    model_limited_preds = predict(model_limited), 
    model_limited_resids = resid(model_limited)
  ) %>% 
  ggplot(aes(model_limited_preds, model_limited_resids)) + 
  geom_point() + 
  stat_smooth()
```
Our comments
- edge effects not lookin too good

###> This doesn't look good. There seems to be a pretty clear non-linear relationship in this data -- perhaps parabolic or logarithmic.
###> To correct this, we will try variable transformations, hoping to end up with a more linear pattern in this plot.
### > Additionally, you may sometimes want to create predictor versus fitted plots, which may give you clues about what variables you need to transform.
```{r}
df <- df %>% 
  mutate(
    model_limited_preds = predict(model_limited), 
    model_limited_resids = resid(model_limited)
  ) 
mask_resids <- df %>% 
  ggplot(aes(mask_ratio, model_limited_resids)) + 
  geom_point() + 
  stat_smooth()
senior_resids <- df %>% 
  ggplot(aes(Senior_Perc, model_limited_resids)) + 
  geom_point() + 
  stat_smooth()
unemployment_resids <- df %>% 
  ggplot(aes(Unemployment_Perc, model_limited_resids)) + 
  geom_point() + 
  stat_smooth()
```

```{r}
mask_resids
senior_resids
unemployment_resids
```
- looks a bit worse, edge effects still


- No perfect collinearity
### > First, we can look at our coefficients, and notice that R has not dropped any variables.
```{r}
model_limited$coefficients
```
- no perfect collinearity

```{r, message = FALSE}
df %>% 
  select(
    Fatality_Rate, mask_ratio, Senior_Perc, Poverty_Perc, model_limited_resids
  ) %>%
  GGally::ggpairs()
```


### > This tells us that there is no perfect collinearity.  This assumption also includes the requirement that a BLP exists, which may not happen if there are heavy tails.  In this case, though, we don't see any distributions that look like they have unusually low or high values.
- Homoskedastic errors
### > To assess whether the distribution of the errors is homoskedastic, we can examine the residuals versus fitted plot again.  We are interested in whether there is a band of even thickness from left to right. Looking at the plot above, indeed, it does look like there might be some increase in the variance of the residuals at the upper side of the predicted values, but it is not severe.  We may hope that a log transform fixes this small problem
### > Another idea is to examine the scale-location plot.  Homoskedasticity would show up on this plot as a flat smoothing curve.
 
```{r}
plot(model_limited, which=3)
```
- heteroskedasticity is prevalent here

### > This plot looks quite good, suggesting no major problem with heteroskedasticity.
- Normally distributed errors
```{r}
plot_one <- df %>% 
  ggplot(aes(x = model_limited_resids)) + 
  geom_histogram()
  
plot_two <- df %>% 
  ggplot(aes(sample = model_limited_resids)) + 
  stat_qq() + stat_qq_line()
plot_one / plot_two
```
- not too bad?? some minor deviation from normality towards the tails
- long right tail
- high frequency of values just below -0.005

> The histogram of residuals and the qqplot shows some some deviation from normality, specifically a right skew and perhaps an unusual concentration on the right tail.
>
> This is not a problem for unbiasedness, and it is not a problem for our standard errors.  However, this will threaten the validity of our t-tests and confidence intervals.  We may hope to fix this problem with a variable transformatin.


### 5. Discussion of Omitted Variables -- All, led by Spencer Song

If the team has taken up an explanatory (i.e. causal) question to evaluate, then identify what you think are the 5 most important *omitted variables* that bias results you care about. For each variable, you should *reason about the direction of bias* caused by omitting this variable. If you can argue whether the bias is large or small, that is even better. State whether you have any variables available that may proxy (even imperfectly) for the omitted variable. Pay particular attention to whether each omitted variable bias is *towards zero or away from zero*. You will use this information to judge whether the effects you find are likely to be real, or whether they might be entirely an artifact of omitted variable bias.
TODO: health
TODO: policy
TODO: socioeconomic
TODO: other - weather, travel zones, travel patterns, political demographics, vaccine distribution, total cases, city density


### 6. Conclusion -- save 4 later

Make sure that you end your report with a discussion that distills key insights from your estimates and addresses your research question.

TODO: all

## Rubric for Evaluation

You may use the following, loosely structured rubric to guide your writing.

- **Introduction.** Is the introduction clear? Is the research question specific and well defined? Does the introduction motivate a specific concept to be measured and explain how it will be operationalized. Does it do a good job of preparing the reader to understand the model specifications?

- **The Initial Data Loading and Cleaning.** Did the team notice any anomalous values? Is there a sufficient justification for any data points that are removed? Did the report note any coding features that affect the meaning of variables (e.g. top-coding or bottom-coding)? Overall, does the report demonstrate a thorough understanding of the data? Does the report convey this understand to its reader -- can the reader, through reading this report, come to the same understanding that the team has come to? 

- **The Model Building Process.** Overall, is each step in the model building process supported by EDA? Is the outcome variable appropriate? Did the team clearly state why they chose these explanatory variables, does this explanation make sense in term of their research question? Did the team consider available variable transformations and select them with an eye towards model plausibility and interpretability? Are transformations used to expose linear relationships in scatterplots? Is there enough explanation in the text to understand the meaning of each visualization?

- **Regression Models:**
   - **Base Model.** Does this model only include key explanatory variables? Do the variables make sense given the measurement goals? Did the team apply reasonable transformations to these variables, to capture the nature of the relationships? Does the team write about this model in prose in a way that is appropriate? 
   - **Second Model.** Does this model represent a balanced approach, including variables that advance modeling goals without causing major issues? Does the model succeed in reducing standard errors of the key variables compared to the base model? Does it capture major non-linearities in the joint distribution of the variables? Does the team write about this model in prose in a way that is appropriate? 
   - **Third Model.** Does this model represent a maximalist approach, erring on the side of including most variables? Is it still a reasonable model? Are there any variables that are outcomes, and should therefore still be excluded? Is there too much colinearity, to the point that the key causal effects cannot be measured? Does this team write about this model in prose in a way that is appropriate? 

- **A Regression Table.** Are the model specifications properly chosen to outline the boundary of reasonable choices? Is it easy to find key coefficients in the regression table? Does the text include a discussion of practical significance for key effects? 
   
- **Plots, Figures, and Tables** Do the plots, figures and tables that the team has chosen to include successfully move forward the argument that they are making? Has the team chosen the most effective method (a table or a chart) to display their evidence? Is that table or chart the most communicative it could be? Is every plot, figure, and table that is included in the report referenced in the narrative argument?

- **Assessment of the CLM.** Has the team presented a sober assessment of the CLM assumptions that might be problematic for their model? Have they presented their analysis about the consequences of these problems (including random sampling) for the models they estimate? Did they use visual tools or statistical tests, as appropriate? Did they respond appropriately to any violations?

- **An Omitted Variables Discussion.** Did the report miss any important sources of omitted variable bias? Are the estimated directions of bias correct? Was their explanation clear? Is the discussion connected to whether the key effects are real or whether they may be solely an artifact of omitted variable bias?

- **Conclusion.** Does the conclusion address the research question? Does it raise interesting points beyond numerical estimates? Does it place relevant context around the results?

- Are there any other errors, faulty logic, unclear or unpersuasive writing, or other elements that leave you less convinced by the conclusions?
