---
title: "EDA of Health Variables"
output: pdf_document
---

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(car)
library(lmtest)
library(sandwich)
library(stargazer)
library(tidyverse)
library(magrittr)
library(ggplot2)
library(dplyr)
library(patchwork)
```


```{r, echo = FALSE, message = FALSE, warning = FALSE}
df <- read_csv('FinalDF.csv')
df <- df[-c(2, 9, 12), ]
```

# Fatality_Rate

```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.height = 3}
fatality_hist <- df %>%
  ggplot() +
  aes(Fatality_Rate) +
  geom_histogram()

fatality_box <- df %>%
  ggplot() +
  aes(Fatality_Rate) +
  geom_boxplot()

fatality_hist | fatality_box

summary(df$Fatality_Rate)
```

`Fatality_Rate` has a relatively normal distribution. There is no significant skew in the data. `Fatality_Rate` ranges from 0.47% to 2.85%. There are 3 outliers.

\pagebreak{}

# Look at health variables used in each model

## Model 1
Population..65..

## Model 2
Population..65.., Diabetes_Diagnosed_Perc, Stroke_Death_Perc, Heart_Disease_Death_Perc,
Smoker_Perc, Chronic_Respiratory_Disease_Death_Perc

## Model 3
Population..65.., Diabetes_Diagnosed_Perc, Stroke_Death_Perc, Heart_Disease_Death_Perc,
Smoker_Perc, Chronic_Respiratory_Disease_Death_Perc, Asthma_Deaths_Perc, Obesity_Perc

\begin{tabular}{ |c|c|c|c| }
 \hline
 Variable & Linear Relationship? & Correlation & Recommendation \\ 
 \hline
 Diabetes & Yes & Positive (0.33) & Include in limited model, no transformation \\  
 Stroke & No & Basically 0 (0.0095) & Don't include at all \\ 
 Heart Disease & Slightly & Positive (0.21) & Include in model 2, no transformation \\ 
 Smokers & Not Really & Positive (0.118) & Include in model 3, no transformation \\ 
 Chronic Resp Dis & Yes & Negative (-0.2239) & Model 2? no transformation \\ 
 Asthma & Slightly & Negative (-0.195) & Model 3 only? no transformation \\ 
 Obesity & Not Really & Basically 0 (0.0348) & Model 3 only? no transformation \\
 \hline
\end{tabular}

# Model 1 EDA
         
```{r model 1 eda, echo = FALSE, message = FALSE, warning = FALSE}
age_hist <- df %>% 
  ggplot() +
  aes(Population..65../Population) +
  geom_histogram()

age_box <- df %>% 
  ggplot() +
  aes(Population..65../Population) +
  geom_boxplot()

age_scatter <- df %>%
  ggplot() +
  aes(Population..65../Population, Fatality_Rate) +
  geom_point() + geom_smooth()

(age_hist | age_box) / age_scatter

summary(df$Population..65../df$Population)
```

The percentage of the population > 65 has a few outliers on each end. It appears that `Fatality_Rate` and % of 65+ have a positive correlation up to a certain point (~17.5% or so). No combination of log transformations makes a more linear relationship between the two variables. % of 65+ is relatively normally distributed. 

# Model 2 EDA
         
## Diabetes_Diagnosed_Perc

```{r model 2 eda - diabetes, echo = FALSE, message = FALSE, warning = FALSE}
diabetes_hist <- df %>% 
  ggplot() +
  aes(Diabetes_Diagnosed_Perc) +
  geom_histogram(bins = 20)

diabetes_box <- df %>%
  ggplot() +
  aes(Diabetes_Diagnosed_Perc) +
  geom_boxplot()

diabetes_scatter <- df %>%
  ggplot() +
  aes(Diabetes_Diagnosed_Perc, Fatality_Rate) +
  geom_point() + geom_smooth()

diabetes_scatter_log <- df %>%
  ggplot() +
  aes((Diabetes_Diagnosed_Perc), sqrt(Fatality_Rate)) +
  geom_point() + geom_smooth()

(diabetes_hist | diabetes_box) / (diabetes_scatter | diabetes_scatter_log)

summary(df$Diabetes_Diagnosed_Perc)
```

```{r}
cor(df$Diabetes_Diagnosed_Perc, df$Fatality_Rate)
```


`Diabetes_Diagnosed_Perc` has a small right tail, though not highly significant, so there is no major deviation from normality. There is one outlier at 11.7%. `Diabetes_Diagnosed_Perc` and `Fatality_Rate` have a positive correlation. Log transformations do not significantly make the relationship between these two variables more linear.

## Stroke_Death_Perc

```{r model 2 eda - stroke, echo = FALSE, message = FALSE, warning = FALSE}
stroke_hist <- df %>% 
  ggplot() +
  aes(Stroke_Death_Perc) +
  geom_histogram(bins = 20)

stroke_box <- df %>%
  ggplot() +
  aes(Stroke_Death_Perc) +
  geom_boxplot()

stroke_scatter <- df %>%
  ggplot() +
  aes(Stroke_Death_Perc, Fatality_Rate) +
  geom_point() + geom_smooth()

stroke_scatter_log <- df %>%
  ggplot() +
  aes(Stroke_Death_Perc, log(Fatality_Rate)) +
  geom_point() + geom_smooth()

(stroke_hist | stroke_box) / (stroke_scatter | stroke_scatter_log)

summary(df$Stroke_Death_Perc)
```

```{r}
cor(df$Stroke_Death_Perc, df$Fatality_Rate)
```


`Stroke_Death_Perc` has a relatively normal distribution. There are no outliers. The `Fatality_Rate` vs `Stroke_Death_Perc` scatterplot has a parabolic shape, but this shape cannot be made more linear using log transformations.

## Heart_Disease_Death_Perc

```{r model 2 eda - heart disease, echo = FALSE, message = FALSE, warning = FALSE}
heart_disease_hist <- df %>% 
  ggplot() +
  aes(Heart_Disease_Death_Perc) +
  geom_histogram(bins = 20)

heart_disease_box <- df %>%
  ggplot() +
  aes(Heart_Disease_Death_Perc) +
  geom_boxplot()

heart_disease_scatter <- df %>%
  ggplot() +
  aes(Heart_Disease_Death_Perc, Fatality_Rate) +
  geom_point() + geom_smooth()

heart_disease_scatter_log <- df %>%
  ggplot() +
  aes(log(Heart_Disease_Death_Perc), Fatality_Rate) +
  geom_point() + geom_smooth()

(heart_disease_hist | heart_disease_box) / (heart_disease_scatter | heart_disease_scatter_log)

summary(df$Heart_Disease_Death_Perc)
```

```{r}
cor(df$Heart_Disease_Death_Perc, df$Fatality_Rate)
```


`Heart_Disease_Death_Perc` appears to be slightly skewed. Data above the median appears to have more spread. There is not a clear positive or negative correlation between `Heart_Disease_Death_Perc` and `Fatality_Rate`. The relationship is relatively flat except for a step between 0.0015 and 0.001625. This variable does not appear to be a good predictor for `Fatality_Rate`

## Smoker_Perc

```{r model 2 eda - smokers, echo = FALSE, message = FALSE, warning = FALSE}
smoker_hist <- df %>% 
  ggplot() +
  aes(Smoker_Perc) +
  geom_histogram(bins = 20)

smoker_box <- df %>%
  ggplot() +
  aes(Smoker_Perc) +
  geom_boxplot()

smoker_scatter <- df %>%
  ggplot() +
  aes(Smoker_Perc, Fatality_Rate) +
  geom_point() + geom_smooth()

smoker_scatter_log <- df %>%
  ggplot() +
  aes(log(Smoker_Perc), Fatality_Rate) +
  geom_point() + geom_smooth()

(smoker_hist | smoker_box) / (smoker_scatter | smoker_scatter_log)

summary(df$Smoker_Perc)
```

```{r}
cor((df$Smoker_Perc), df$Fatality_Rate)
```


`Smoker_Perc` has a relative normal distribution, but there is no clear linear relationship between `Smoker_Perc` and `Fatality_Rate`. Applying a log transformation does not make the relationship any more linear.

## Chronic_Respiratory_Disease_Death_Perc

```{r model 2 eda - chronic respiratory disease, echo = FALSE, message = FALSE, warning = FALSE}
chron_resp_dis_hist <- df %>% 
  ggplot() +
  aes(Chronic_Respiratory_Disease_Death_Perc) +
  geom_histogram(bins = 20)

chron_resp_dis_box <- df %>%
  ggplot() +
  aes(Chronic_Respiratory_Disease_Death_Perc) +
  geom_boxplot()

chron_resp_dis_scatter <- df %>%
  ggplot() +
  aes(Chronic_Respiratory_Disease_Death_Perc, Fatality_Rate) +
  geom_point() + geom_smooth()

chron_resp_dis_scatter_log <- df %>%
  ggplot() +
  aes(log(Chronic_Respiratory_Disease_Death_Perc), Fatality_Rate) +
  geom_point() + geom_smooth()

(chron_resp_dis_hist | chron_resp_dis_box) / (chron_resp_dis_scatter | chron_resp_dis_scatter_log)

summary(df$Chronic_Respiratory_Disease_Death_Perc)
```

```{r}
cor(df$Chronic_Respiratory_Disease_Death_Perc, df$Fatality_Rate)
```


Pre-Removing Outliers:

`Chronic_Respiratory_Disease_Death_Perc` has a relatively normal distribution and does not have any outliers. There is not a clear linear relationship between `Chronic_Respiratory_Disease_Death_Perc` and `Fatality_Rate`. There appears to be some parabolic relationship when `Chronic_Respiratory_Disease_Death_Perc` is less than 5e-4, but greater than this number, there is no relationship.

Post-Removing Outliers:

Overall negative relationship between `Chronic_Respiratory_Disease_Death_Perc` and `Fatality_Rate`

# Model 3

## Asthma_Deaths_Perc
         
```{r model 3 eda - asthma, echo = FALSE, message = FALSE, warning = FALSE}
asthma_hist <- df %>% 
  ggplot() +
  aes(Asthma_Deaths_Perc) +
  geom_histogram(bins = 20)

asthma_box <- df %>%
  ggplot() +
  aes(Asthma_Deaths_Perc) +
  geom_boxplot()

asthma_scatter <- df %>%
  ggplot() +
  aes(Asthma_Deaths_Perc, Fatality_Rate) +
  geom_point() + geom_smooth()

asthma_scatter_log <- df %>%
  ggplot() +
  aes(log(Asthma_Deaths_Perc), Fatality_Rate) +
  geom_point() + geom_smooth()

(asthma_hist | asthma_box) / (asthma_scatter | asthma_scatter_log)

summary(df$Asthma_Deaths_Perc)
```

```{r}
cor((df$Asthma_Deaths_Perc), df$Fatality_Rate)
```


`Asthma_Deaths_Perc` has a few outliers at each of its tails. The distribution is not quite normal. There appears to be somewhat of a linear relationship between `Fatality_Rate` and `Asthma_Deaths_Perc`, but if anything, it looks like the correlation is negative, which does not support our intuition. Applying a log transformation to `Asthma_Deaths_Perc` does not help make the relationship more linear.

## Obesity_Perc

```{r model 3 eda - obesity, echo = FALSE, message = FALSE, warning = FALSE}
obesity_hist <- df %>% 
  ggplot() +
  aes(Obesity_Perc) +
  geom_histogram(bins = 20)

obesity_box <- df %>%
  ggplot() +
  aes(Obesity_Perc) +
  geom_boxplot()

obesity_scatter <- df %>%
  ggplot() +
  aes(Obesity_Perc, Fatality_Rate) +
  geom_point() + geom_smooth()

obesity_scatter_log <- df %>%
  ggplot() +
  aes(log(Obesity_Perc), Fatality_Rate) +
  geom_point() + geom_smooth()

(obesity_hist | obesity_box) / (obesity_scatter | obesity_scatter_log)

summary(df$Obesity_Perc)
```

```{r}
cor((df$Obesity_Perc), df$Fatality_Rate)
```


`Obesity_Perc` does not have any outliers, but it does not appear to follow a normal distribution either. The relationship between `Obesity_Perc` and `Fatality_Rate` appears quadratic or cubic (especially when a log transformation is applied to `Obesity_Perc`). Applying a transformation does not make the data appear more linear.


Population..65.., Diabetes_Diagnosed_Perc, Stroke_Death_Perc, Heart_Disease_Death_Perc,
Smoker_Perc, Chronic_Respiratory_Disease_Death_Perc, Asthma_Deaths_Perc, Obesity_Perc

```{r}
model <- lm(Fatality_Rate ~ Diabetes_Diagnosed_Perc + Chronic_Respiratory_Disease_Death_Perc + Obesity_Perc, data = df)
coeftest(model)
# + Stroke_Death_Perc + Heart_Disease_Death_Perc + Smoker_Perc + 
# + Asthma_Deaths_Perc
```

```{r}
df %>% 
  mutate(
    model_preds = predict(model), 
    model_resids = resid(model)
  ) %>% 
  ggplot(aes(model_preds, model_resids)) + 
  geom_point() + 
  stat_smooth()
```