---
title: "load_data"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
install.packages("googlesheets")
library(gsheet)
library(readr)
library(readxl)
library(tidyverse)
library(car)
library(lmtest)
library(sandwich)
library(stargazer)
library(tidyverse)
library(magrittr)
library(ggplot2)
library(dplyr)
library(patchwork)
```

```{r}
# Load county-level COVID case and death data from NY times
url_nytimes <- 'https://raw.githubusercontent.com/nytimes/covid-19-data/master/live/us-counties.csv'
cases_deaths <- read_csv(url(url_nytimes))

# Load county-level data about icu beds from CovidCareMap (CCM)
url_covidcaremap <- 'https://raw.githubusercontent.com/covidcaremap/covid19-healthsystemcapacity/master/data/published/us_healthcare_capacity-county-CovidCareMap.csv'
icu_beds_ccm <- read_csv(url(url_covidcaremap))
```


```{r}
cases_deaths_edit <-
  cases_deaths %>%
    select(county, state, fips, cases, deaths) %>%
    transform(fips = as.numeric(fips)) %>%
    transform(fips = as.character(fips))
```

```{r}
icu_beds_ccm_edit <-
  icu_beds_ccm %>%
    select(fips_code, `County Name`, `Staffed All Beds`, `Staffed ICU Beds`, `Staffed ICU Beds [Per 1000 Elderly (65+)]`, `Staffed ICU Beds [Per 1000 Adults (20+)]`, `Staffed ICU Beds [Per 1000 People]`, Population, `Population (65+)`, `Population (20+)`) %>%
    transform(fips_code = as.numeric(fips_code)) %>%
    transform(fips_code = as.character(fips_code))
```

```{r}
final_f <- merge(x = cases_deaths_edit, y = icu_beds_ccm_edit, by.x = "fips", by.y = "fips_code", all.x = TRUE)
final_f <- 
  final_f %>%
    mutate(perc_65 = Population / Population..65..)
```

#spencer hong
```{r}
# Load data about mask-wearing polices from the W203 spreadsheet
url_203data_masks <- 'https://docs.google.com/spreadsheets/d/1zu9qEWI8PsOI_i8nI_S29HDGHlIp2lfVMsGxpQ5tvAQ/edit#gid=1489353670'
mask_policy <- read.csv(text=gsheet2text(url_203data_masks,format='csv'),
                        stringsAsFactors=FALSE)
```

```{r}
# Load data about stay-at-home orders from the W203 spreadsheet
url_203data_stayathome <- 'https://docs.google.com/spreadsheets/d/1zu9qEWI8PsOI_i8nI_S29HDGHlIp2lfVMsGxpQ5tvAQ/edit#gid=1894978869'
stayathome_policy <- read.csv(text=gsheet2text(url_203data_stayathome,format='csv'),
                        stringsAsFactors=FALSE)
```

## Prepare Mask Policy Data
#Calculate the longest number of days that any state had a mask mandate policy, which would be whichever state implemented it first and then up to the most recent date a state ended their mask mandate
-Column 'Public face mask mandate' = beginning date
-Column 'End face mask mandate' = ending date
```{r, Calculate mask mandate date range}
mask_policy <- mask_policy[-52,]

first_mask <- mask_policy %>% filter(mask_policy$Public.face.mask.mandate != 0)
last_mask <- mask_policy %>% filter(mask_policy$End.face.mask.mandate != 0)

first_mask$Public.face.mask.mandate <- as.Date(first_mask$Public.face.mask.mandate, format = "%m/%d/%Y")
last_mask$End.face.mask.mandate <- as.Date(last_mask$End.face.mask.mandate, format = "%m/%d/%Y")

mask_begin <- min(first_mask$Public.face.mask.mandate)
mask_end <- max(last_mask$End.face.mask.mandate)

mask_range <- mask_end - mask_begin

mask_range <- as.numeric(mask_range, units = "days")
```

#Add a column for the number of days each state had/has a mask mandate policy
```{r, Calculate number of days each state had a mask policy}
mask_policy$Public.face.mask.mandate <- as.Date(mask_policy$Public.face.mask.mandate, format = "%m/%d/%Y")
mask_policy$End.face.mask.mandate <- as.Date(mask_policy$End.face.mask.mandate, format = "%m/%d/%Y")

mask_policy <- mask_policy %>% mutate(days_mask = case_when(
  is.na(Public.face.mask.mandate) == TRUE ~ 0,
  is.na(Public.face.mask.mandate) == FALSE & is.na(End.face.mask.mandate) == TRUE ~ as.numeric((mask_end - Public.face.mask.mandate), units = "days"),
  TRUE ~ as.numeric((End.face.mask.mandate - Public.face.mask.mandate), units = "days")))
```

#Add a column to the mask_policy data frame to show the ratio of days each state mandate masks
```{r, Calculate ratio of policy in place to the total number of days masks have been required}
mask_policy$mask_ratio <- mask_policy$days_mask / mask_range
```

## Prepare Stay-at-Home Policy Data
#Calculate the longest number of days that any state had a stay-at-home mandate policy, which would be whichever state implemented it first and then up to the most recent date a state ended their stay-at-home mandate
-Column 'Stay at home/shelter in place' = beginning date
-Column 'End stay at home/shelter in place ' = ending date
```{r, Calculate stay-at-home mandate date range}
stayathome_policy <- stayathome_policy[-52,]

first_stay <- stayathome_policy %>% filter(stayathome_policy$Stay.at.home.shelter.in.place != 0)
last_stay <- stayathome_policy %>% filter(stayathome_policy$End.stay.at.home.shelter.in.place != 0)

first_stay$Stay.at.home.shelter.in.place <- as.Date(first_stay$Stay.at.home.shelter.in.place, format = "%m/%d/%Y")
last_stay$End.stay.at.home.shelter.in.place <- as.Date(last_stay$End.stay.at.home.shelter.in.place, format = "%m/%d/%Y")

home_begin <- min(first_stay$Stay.at.home.shelter.in.place)
home_end <- max(last_stay$End.stay.at.home.shelter.in.place)

home_range <- home_end - home_begin

home_range <- as.numeric(home_range, units = "days")
```

#Add a column for the number of days each state had/has a stay-at-home policy
```{r, Calculate number of days each state had a mask policy}
stayathome_policy$Stay.at.home.shelter.in.place <- as.Date(stayathome_policy$Stay.at.home.shelter.in.place, format = "%m/%d/%Y")
stayathome_policy$End.stay.at.home.shelter.in.place <- as.Date(stayathome_policy$End.stay.at.home.shelter.in.place, format = "%m/%d/%Y")

stayathome_policy <- stayathome_policy %>% mutate(days_home = case_when(
  is.na(Stay.at.home.shelter.in.place) == TRUE ~ 0,
  is.na(Stay.at.home.shelter.in.place) == FALSE & is.na(End.stay.at.home.shelter.in.place) == TRUE ~ as.numeric((home_end - Stay.at.home.shelter.in.place), units = "days"),
  TRUE ~ as.numeric((End.stay.at.home.shelter.in.place - Stay.at.home.shelter.in.place), units = "days")))
```

#Add a column to the stayathome_policy data frame to show the ratio of days each state mandate to stay at home 
```{r, Calculate ratio of policy in place to the total number of days masks have been required}
stayathome_policy$home_ratio <- stayathome_policy$days_home / home_range
```

## Merge Data Frames
#Merge mask and stay-at-home policy data sets
``` {r, Merge data frames}
policy_data <- merge(mask_policy, stayathome_policy, by.x="State.FIPS.Code", by.y="State.FIPS.Code")
```

#Remove columns
``` {r, Remove columns from new data frame}
policy_data <- policy_data[-c(4:13, 16:23)]
```

#Remove columns
``` {r, Remove columns from new data frame}
policy_data <- policy_data[-c(4,6)]
```


```{r}
# Load data from Yu-Chronic Respiratory Disease Mortality
# There are several sheets in this xlsx file. Relevant sheets/variables to pull:
# 1. Chronic respiratory diseases

chronic_respiratory_diesease <- read_excel('data/Yu-Chronic Respiratory Disease Mortality/IHME_USA_COUNTY_RESP_DISEASE_MORTALITY_1980_2014_NATIONAL_Y2017M09D26.XLSX',
                                           sheet = 'Chronic respiratory diseases')

asthma_death <- read_excel('data/Yu-Chronic Respiratory Disease Mortality/IHME_USA_COUNTY_RESP_DISEASE_MORTALITY_1980_2014_NATIONAL_Y2017M09D26.XLSX',
                           sheet = 'Asthma')
```


```{r}
# Load data from Yu-County Health Rankings
# Variables in dataset (confirm in documentation):
# Adult smoking (% smokers)
# Adult obesity (% adults with obesity)
# % 65 and older

county_health_rankings <- read_csv('data/Yu-County Health Rankings/analytic_data2021.csv')
```


```{r}
# Load diabetes data from Yu-Diabetes
# Diagnosed Diabetes; Total (Total); Adults Aged 20+ Years; Crude Number; U.S. Counties; 2017

# Edited the csv so that the column headers are in the top row
diabetes <- read_csv('data/Yu-Diabetes/DiabetesAtlasCountyData-Edited.csv')
```

```{r}
# Load data from Yu-Stroke and Heart Disease
heart_disease <- read_csv('data/Yu-Stroke and Heart Disease/all_heart_disease_deathrateper100k_allages_2016-2018.csv')
stroke <- read_csv('data/Yu-Stroke and Heart Disease/all_stroke_deathrateper100k_allages_2016-2018.csv')
```

```{r astha handling}
# Get asthma death data in usable format

# Select relevant columns: Location, FIPS, and 2014 mortality rate
asthma_death_2014_clean <-  asthma_death %>%
  select(Asthma, ...2, ...10)

# Rename columns and convert FIPS to integer
# take average mortality rate (delete confidence interval)
asthma_death_2014_clean <- asthma_death_2014_clean %>% 
  rename(
    Location = Asthma,
    FIPS = ...2,
    Mortality_Rate_2014 = ...10
    ) %>%
  transform(
    FIPS = as.numeric(FIPS),
    Mortality_Rate_2014 = as.numeric(substr(Mortality_Rate_2014, start = 1, stop = 4))
  )

# Remove top row
asthma_death_2014_clean <- asthma_death_2014_clean[-c(1), ]


summary(asthma_death_2014_clean)
```

```{r chronic respiratory disease handling}
# Get chronic respiratory disease death data in usable format

# Select relevant columns: Location, FIPS, and 2014 mortality rate
chronic_respiratory_diesease_2014_clean <-  chronic_respiratory_diesease %>%
  select("Chronic respiratory diseases", ...2, ...10)

# Rename columns and convert FIPS to integer
# take average mortality rate (delete confidence interval)
chronic_respiratory_diesease_2014_clean <- chronic_respiratory_diesease_2014_clean %>%
  rename(
    Location = "Chronic respiratory diseases",
    FIPS = ...2,
    Mortality_Rate_2014 = ...10
    ) %>%
  transform(
    FIPS = as.numeric(FIPS),
    Mortality_Rate_2014 = as.numeric(substr(Mortality_Rate_2014, start = 1, stop = 4))
  )

# Remove top row
chronic_respiratory_diesease_2014_clean <- chronic_respiratory_diesease_2014_clean[-c(1), ]


summary(chronic_respiratory_diesease_2014_clean)
```



```{r}
# Variables:
# 1. % of adults with obesity
# 2. % of adults that smoke
# 3. % of adults that are 65+
county_health_rankings_clean <- county_health_rankings %>%
  select(
    `5-digit FIPS Code`,
    `Adult obesity raw value`,
    `Adult smoking raw value`, 
    `% 65 and older raw value`,
  ) %>%
  rename(
    FIPS = `5-digit FIPS Code`,
    obesity = `Adult obesity raw value`,
    smoking = `Adult smoking raw value`,
    over_65 = `% 65 and older raw value`
    ) %>%
  transform(
    FIPS = as.numeric(FIPS),
    obesity = as.numeric(obesity),
    smoking = as.numeric(smoking),
    over_65 = as.numeric(over_65)
  )
```

```{r diabetes cleaning}
# people diagnosed with diabetes
# needs to be normalized

# select appropriate columns
# convert to numbers
diabetes_clean <- diabetes %>%
  select(
    CountyFIPS,
    Number
  ) %>%
  rename(
    FIPS = CountyFIPS,
    num_diabetes_diagnosed = Number
  ) %>%
  transform(
    FIPS = as.numeric(FIPS),
    num_diabetes_diagnosed = as.numeric(num_diabetes_diagnosed)
  )

summary(diabetes_clean)
```


```{r heart disease}
# heart disease per 100k
# -1 values = NA
# already normalized

# select appropriate columns
# convert to numbers
heart_disease_clean <- heart_disease %>%
  select(
    cnty_fips,
    Value
  ) %>%
  rename(
    FIPS = cnty_fips,
    heart_disease_per100k = Value
  ) %>%
  transform(
    FIPS = as.numeric(FIPS),
    heart_disease_per100k = ifelse(as.numeric(heart_disease_per100k) > -1, as.numeric(heart_disease_per100k), NA)
  )

summary(heart_disease_clean)
```

```{r stroke}
# stroke per 100k
# -1 values = NA
# already normalized

# select appropriate columns
# convert to numbers
stroke_clean <- stroke %>%
  select(
    cnty_fips,
    Value
  ) %>%
  rename(
    FIPS = cnty_fips,
    stroke_per100k = Value
  ) %>%
  transform(
    FIPS = as.numeric(FIPS),
    stroke_per100k = ifelse(as.numeric(stroke_per100k) > -1, as.numeric(stroke_per100k), NA)
  )

summary(stroke_clean)
```


```{r produce final df}
all_health_data_clean <- diabetes_clean %>% merge(asthma_death_2014_clean, by = "FIPS", all.x = TRUE) %>%
  merge(chronic_respiratory_diesease_2014_clean, by = "FIPS", all.x = TRUE) %>%
  merge(county_health_rankings_clean, by = "FIPS", all.x = TRUE) %>%
  merge(heart_disease_clean, by = "FIPS", all.x = TRUE) %>%
  merge(stroke_clean, by = "FIPS", all.x = TRUE)
```



#kevin lu
#Extract SVI Data Columns
#Below Poverty
#Unemployed
#Income
#Aged 65 or older
#Minority
#Multi-Unit Structures
#Crowding
```{r} 
# Load data from Yu-Social Vulnerability Index
# svi = social vulnerability index
svi <-  read_csv('data/Yu-Social Vulnerability Index/SVI2018_US_COUNTY.csv')
svi_clean <- data.frame(svi$FIPS, svi$EP_POV/100, svi$EP_UNEMP/100, svi$E_PCI, svi$EP_AGE65/100, svi$EP_MINRTY/100, svi$EP_CROWD/100, svi$EP_MUNIT/100)
svi_clean <- svi_clean %>%
  rename(
    fips_code = svi.FIPS,
    poverty_pct = svi.EP_POV.100,
    unemployment_pct = svi.EP_UNEMP.100,
    income = svi.E_PCI,
    age_65_pct = svi.EP_AGE65.100,
    minority_pct = svi.EP_MINRTY.100,
    multi_unit_pct = svi.EP_MUNIT.100,
    crowding_pct = svi.EP_CROWD.100
    )
svi_clean[, 'fips_code'] <- as.character(as.numeric(svi_clean[, 'fips_code']))
svi_clean <- svi_clean[svi_clean$poverty_pct > 0,]
#summary(svi_clean)
```

```{r}
##**TODO filter out territories, check for more duplicates, check summary for excessive NAs and removing extra columns ***

#final_f(3246), all_health_data_clean(3235), policy_data (51), svi_clean (3141)
merge1 <- merge(x = final_f, y = all_health_data_clean, by.x = "fips", by.y = "FIPS")#, all.x = TRUE)
merge1_clean <- merge1 %>%
  filter(is.na(fips) == FALSE)
merge2 <- merge(x = merge1_clean, y = svi_clean, by.x = "fips", by.y = "fips_code", all.x = TRUE)
merge3 <- merge(x = merge2, y = policy_data, by.x = "state", by.y = "State.x", all.x = TRUE)
#merge_dups <- merge1[duplicated(merge1$fips),]

print(colnames(merge3))
# "state"                                      "fips"*                                     
#  [3] "county"                                    "cases"                                    
#  [5] "deaths" *normalize death rate*                                   "County.Name"    *                          
#  [7] "Staffed.All.Beds"                          "Staffed.ICU.Beds"                         
#  [9] "Staffed.ICU.Beds..Per.1000.Elderly..65..." "Staffed.ICU.Beds..Per.1000.Adults..20..." 
# [11] "Staffed.ICU.Beds..Per.1000.People."        "Population"                               
# [13] "Population..65.."                          "Population..20.."                         
# [15] "perc_65"                                   "num_diabetes_diagnosed" (normalized)                   
# [17] "Location.x" *                              "Mortality_Rate_2014.x"                    
# [19] "Location.y" *                              "Mortality_Rate_2014.y"                    
# [21] "obesity"                                   "smoking"                                  
# [23] "over_65"                                   "heart_disease_per100k"
# [25] "stroke_per100k"                            "poverty_pct"
# [27] "unemployment_pct"                          "income"                                   
# [29] "age_65_pct"   *                             "minority_pct"                             
# [31] "crowding_pct"                              "multi_unit_pct"                           
# [33] "State.FIPS.Code"*                           "State.Abbreviation.x"*                     
# [35] "mask_ratio"                                "home_ratio"      

#merge2
summary(merge2)

#merge(x = final_r, y = icu_beds_ccm_edit, by.x = "fips", by.y = "fips_code")
```




