---
title: "EDA of Policy Variables"
output: html_document
---

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(car)
library(lmtest)
library(sandwich)
library(stargazer)
library(tidyverse)
library(magrittr)
library(ggplot2)
library(dplyr)
library(patchwork)
library(olsrr)
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
df <- read_csv('FinalDF.csv')
```

# Look at policy variables used in each model

## Model 1
mask_ratio

## Model 2
mask_ratio, home_ratio

## Model 3 - Same as Model 2
mask_ratio, home_ratio

# Model 1 EDA
## mask_ratio
         
```{r Model 1 EDA - Mask Mandate, echo = FALSE, message = FALSE, warning = FALSE, fig.height = 3}
mask_hist <- df %>%
  ggplot() +
  aes(mask_ratio) +
  geom_histogram()

mask_box <- df %>%
  ggplot() +
  aes(mask_ratio) +
  geom_boxplot()

mask_scatter <- df %>%
  ggplot() +
  aes(mask_ratio, Fatality_Rate) +
  geom_point() + geom_smooth()

(mask_hist | mask_box) / mask_scatter

summary(df$mask_ratio)

cor(df$mask_ratio, df$Fatality_Rate)
```

There are no outliers for the percentage duration that states enacted mask mandate policies. It appears that `Fatality_Rate` and percentage duration of mask policies of states has a positive correlation. No combination of log transformations makes a more linear relationship between the two variables. `mask_ratio` has a non-normal distribution with two modal groups of states. One is for states that implemented a mask mandate for 25% or less of the total duration time while the other group of states implemented mask-wearing for more than 60% of the duration. Given that 13 states did not implement a mask policy there is a left-tail skewness, which is confirmed by the larger median than mean.

# Model 2 EDA
## home_ratio

```{r Model 2 EDA - Stay-at-Home Mandate, echo = FALSE, message = FALSE, warning = FALSE}
home_hist <- df %>%
  ggplot() +
  aes(home_ratio) +
  geom_histogram()

home_box <- df %>%
  ggplot() +
  aes(home_ratio) +
  geom_boxplot()

home_scatter <- df %>%
  ggplot() +
  aes(home_ratio, Fatality_Rate) +
  geom_point() + geom_smooth()

(home_hist | home_box) / home_scatter

summary(df$home_ratio)

cor(df$home_ratio, df$Fatality_Rate)
```

New Mexico and California represent the two outliers for the percentage duration that states enacted stay-at-home policies. New Mexico never ended its policy while California had the longest tenured policy representing 100%. It appears that `Fatality_Rate` and percentage duration of stay-at-home policies of states has a positive correlation until 0.20. No combination of log transformations makes a more linear relationship between the two variables. Without the two outliers, `home_ratio` has a normal distribution with no significant skewness and ranging from 0.0% to 100.0%.

```{r Covariance, echo = FALSE, message = FALSE, warning = FALSE}
cov(df$mask_ratio, df$Fatality_Rate)

cov(df$home_ratio, df$Fatality_Rate)
```

The covariance for both the duration of a mask mandate and the duration of a stay-at-home policy to fatality rate are positive, albeit quite small as both are less than 0.0008. The positive covariance metrics indicate a positive relationship, which is concerning as we assumed they would be negative as longer mask and stay-at-home policies were used with the idea that they would curb infections and therefore deaths.

```{r Multi-Collinearity Check, echo = FALSE, message = FALSE, warning = FALSE}
vif_model <- lm(Fatality_Rate ~ mask_ratio + home_ratio, data = df)

ols_vif_tol(vif_model)
```
The low VIF values indicate that the mask and stay-at-home policy variables are not highly collinear with one another.